---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
### Unzipping, if data file not already present
```{r, results='markup', warning=TRUE, echo=TRUE}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
```
### Loading the dataset from file
```{r, results='markup', warning=TRUE, echo=TRUE}
activityData <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?
### Histogram of the total number of steps taken each day
```{r, results='markup', warning=TRUE, echo=TRUE}
stepsPerDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
hist(stepsPerDay, breaks=seq(from=0, to=25000, by=1000),
     main='Histogram: Total number of steps taken each day',
     xlab='Total steps/day', 
     ylab='Frequency',
     col="blue")
```

### Mean and median number of steps taken each day
```{r, results='markup', warning=TRUE, echo=TRUE}
stepsPerDayMean   <- mean(stepsPerDay)
stepsPerDayMedian <- median(stepsPerDay)
```
* Mean:    `r stepsPerDayMean`
* Median:  `r stepsPerDayMedian`

## What is the average daily activity pattern?
```{r, results='markup', warning=TRUE, echo=TRUE}
avgStepsPerInterval <- aggregate(x=list(meanSteps = activityData$steps),
                                 by=list(interval  = activityData$interval),
                                 FUN=mean, na.rm=TRUE)
library(ggplot2)
ggplot(data=avgStepsPerInterval, aes(x=interval, y=meanSteps)) +
  geom_line() +
  xlab("interval") +
  ylab("average steps/day")  
```

### The 5-minute interval that, on average, contains the maximum number of steps
```{r, results='markup', warning=TRUE, echo=TRUE}
maxStepIndex   <- which.max(avgStepsPerInterval$meanSteps)
maxStepTimeRaw <- avgStepsPerInterval[maxStepIndex,'interval']
maxStepTimeFormatted <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", maxStepTimeRaw)
```
* Interval = `r  maxStepTimeFormatted`

## Imputing missing values
Missing values are filled with the average of the corresponding interval
```{r, results='markup', warning=TRUE, echo=TRUE}
imputedData  <- activityData
invalidIndex <- is.na(imputedData$steps)
mergedData   <- merge(imputedData[invalidIndex, ], avgStepsPerInterval, by = "interval")
imputedData[invalidIndex,]$steps <- mergedData$meanSteps
```
*Number of obervations with missing values: `r sum(invalidIndex)`

### Histogram of the total number of steps taken each day after missing values are imputed
```{r, results='markup', warning=TRUE, echo=TRUE}
imputedStepsPerDay <- tapply(imputedData$steps, imputedData$date, sum)
hist(imputedStepsPerDay, breaks=seq(from=0, to=25000, by=1000),
     main='Histogram: Total number of steps taken each day (Missing Data imputed)',
     xlab='Total steps/day', 
     ylab='Frequency',
     col="blue")
```

## Are there differences in activity patterns between weekdays and weekends?
```{r, results='markup', warning=TRUE, echo=TRUE}
imputedData$dayType <- ifelse(as.POSIXlt(imputedData$date)$wday %in% c(0,6), 'Weekend', 'Weekday')
avgImputedbyDayType <- aggregate(steps ~ interval + dayType, data=imputedData, mean)
ggplot(avgImputedbyDayType, aes(interval, steps)) + 
  geom_line() + 
  facet_grid(dayType ~ .) +
  xlab("interval") + 
  ylab("average(steps)")
```
